Polymethylation scores paper

### 2021-10-26 Papers/Polymethylation_scores/04_background.Rmd
wrote discussion and methods

---
output:
  word_document: default
  html_document: default
---
# Background

Maternal prenatal smoking is a public health concern: in 2018, 11% of pregnant US women aged 15-44 reported any past-month cigarette use, and exposed children are more likely to have low-birth weight, negative neurodevelopmental outcomes and asthma. Measuring prenatal maternal smoking behavior is challenging.  Maternal prenatal smoking is underreported due to stigma. The gold-standard for smoking measurements, serum cotinine, has a short half-life of nine hours in pregnant women.   Further, serum cotinine measures during pregnancy are rarely available outside of birth cohorts, limiting clinical application as a marker of prenatal exposures. Developing a persistent, reliable and feasible biomarker of prenatal maternal smoke exposure is critical for advancing research and precision health.

DNA methylation has been associated with prenatal maternal smoking at various cross-sectional time points across the life course in primarily European ancestry groups. Cordblood, peripheral blood and placental DNA methylation has been significantly associated with prenatal maternal smoking. [Joubert 2016; Bustamante 2021; Davey Smith 2018]. Blood DNA methylation signatures in cohorts of older children and adults also were significantly associated with prenatal maternal smoking exposure [Terry 2019, Davey Smith 2018]. Yet few studies have examined associations at different ages in the same participants. Additionally, while associations are well-replicated in cohorts of European ancestry, few studies have used non-European ancestry populations and those that have had small sample sizes: Fallin 2016 186 individuals of African or admixed ancestry, Terry 2018 60% African or hispanic but only 89 participants, Burchard 2019 572 Latino children]. DNA methylation is associated with prenatal maternal smoking, but the consistency of this association across age and ancestry is understudied.

The association between prenatal maternal smoking and DNA methylation suggests DNA methylation could provide a reliable and specific biomarker of prenatal maternal smoke exposure. Importantly, changes in DNA methylation are specific to smoke exposure *in utero*. Previous work has found reasonable performance of cord blood based [London 2017] (accuracy 91%, sens 58%, spec 97%) [Fallin 2016] (81% accuracy) and peripheral blood based biomarkers (AUC: 0.72) [Davey Smith 2018] of prenatal maternal smoking.A 2018 study found that DNA methylation based scores could even predict predict prenatal smoke exposure (30 years previously) with an area under the curve of 0.72 (95% confidence interval 0.69, 0.76) [Davey Smith 2018]. However, the consistency of DNA methylation as a biomarker for prenatal maternal smoking across age and ancestry has not been evalutated. Few studies have compared the effectiveness of DNA methylation as biomarker of prenatal maternal smoking in the same individuals at different times or evaluated if the inclusion of multiple time points improves performance. Additionally, previous work has relied on blood and placental samples. Cord blood or placenta samples are unlikely to be available in a clinical setting. While peripheral blood is a more realistic clinical sample,  saliva is even easier to collect. To our knowledge, no study has evaluated salivary DNA methylation as a biomarker for prenatal maternal smoking.

>>>>

---
output:
  word_document: default
  html_document: default
bibliography: Methylation_SmokingBib.bib
---

# Background

Maternal prenatal smoking is a public health concern: in 2018, 11% of pregnant US women aged 15-44 reported any past-month cigarette use [@SAMHSA2018], and exposed children are more likely to have low-birth weight, negative neurodevelopmental outcomes and asthma [@Banderali:2015vu]. Measuring prenatal maternal smoking behavior is challenging. Maternal prenatal smoking is underreported due to stigma [@Kvalvik:2012tk; @Dietz:2011ua]. The gold-standard for smoking measurements, serum cotinine, has a short half-life of nine hours in pregnant women [@Dempsey:2002wr]. Further, serum cotinine measures during pregnancy are rarely available outside of birth cohorts, limiting clinical application as a marker of prenatal exposures. Developing a portable and reliable biomarker of prenatal maternal smoke exposure would therefore have important implications for research and clinical practice.

DNA methylation has been associated with prenatal maternal smoking at various cross-sectional time points across the life course in primarily European ancestry groups. Cordblood, peripheral blood and placental DNA methylation has been significantly associated with prenatal maternal smoking [@Joubert:2016ue; @Everson:2021td; @Richmond:2018tu; @Cardenas2019]. Blood DNA methylation signatures in cohorts of older children and adults also were significantly associated with prenatal maternal smoking exposure [@Tehranifar:2018tb ; @Richmond:2018tu]. Yet few studies have examined associations at different ages in the same participants. Additionally, while associations are well-replicated in cohorts of European ancestry, only a few, small sample size studies are in non-European ancestry populations. A 2016 study with 186 participants of African or admixed ancestry [@Ladd-Acosta:2016aa], while a 2018 study of 89 participants had 60% of African or admized ancestry [@Tehranifar:2018tb]. A larger 2019 study of 572 Latino children replicated associations with prenatal maternal smoking at 7 out of 148 tested *a priori* CpG sites from Joubert et al. at a FDR-adjusted *P*-value\<0.05 [@Neophytou:2019ty; @Joubert:2016ue]. Thus, while DNA methylation is associated with prenatal maternal smoking, the consistency of this association across age and ancestry is understudied.

The association between prenatal maternal smoking and DNA methylation suggests DNA methylation could provide a reliable and specific biomarker of prenatal maternal smoke exposure. Importantly, changes in DNA methylation are specific to smoke exposure *in utero*. Previous work has found reasonable performance of cord-blood accuracy ranging from 81% to 91%) [@Reese:2017ts; @Ladd-Acosta:2016aa]. A 2018 study found that DNA methylation based scores from adult peripheral blood could even predict predict prenatal smoke exposure (30 years previously) with an area under the curve of 0.72 (95% confidence interval 0.69, 0.76) [@Richmond:2018tu]. However, the consistency of DNA methylation as a biomarker for prenatal maternal smoking across age and ancestry has not been evaluated. Few studies have compared the effectiveness of DNA methylation as biomarker of prenatal maternal smoking in the same individuals at different times or evaluated if the inclusion of multiple time points improves performance. Additionally, previous work has relied on blood and placental samples. DNA methylation varies across tissues, an epigenetic markers of prenatal maternal smoking are known to differ between cord blood and placental samples[@Everson:2021td]. Cord blood or placenta samples are unlikely to be available in a clinical setting. While peripheral blood is a more realistic clinical sample, saliva is even easier to collect. To our knowledge, no study has evaluated salivary DNA methylation as a biomarker for prenatal maternal smoking.

### 2021-10-26 Papers/Polymethylation_scores/index_paper.Rmd
wrote discussion and methods

---
title: "Polymethylation scores paper"
author: "F blostein"
date: "9/13/2021"
output: pdf_document
bib: Methylation_SmokingBib.bib
---

@@

>>>>

---
title: "Polymethylation scores paper"
author: "F blostein"
date: "9/13/2021"
output:
  word_document: default
  html_document:
    df_print: paged
bibliography: Methylation_SmokingBib.bib
csl: springer-vancouver.csl
---

@@

```{r child = here::here("Papers", "Polymethylation_scores", "08_methods.Rmd")}
```
